Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations

### 2022-02-02 document/source/body.tex
final det paper edits

\section{Preliminaries} \label{s:prelim}
\subsection{Definitions}
Several definitions are introduced in this section. These are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
\\

@@

A simulation can be thought of as the process of generating or producing experimental data. In the case of a driving simulator, kinematics will describe future states of actors given the current conditions and actions taken, thereby generating new data. If a simulation is deterministic, Fig.~\ref{variance_description_b}, then there will be no variation in the generated output data, i.e. all future states are perfectly reproducible from prior states and actions. However, if a simulation is non-deterministic, Fig.~\ref{variance_description_a}, then there will be a variation in the generated output data. \\

@@

Investigating the determinism of game engines has not attracted much research interest to date since performance is more critical for game developers than accurate and repeatable execution. Ensuring software operates deterministically is a non-trivial task.  Catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}. This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.

@@

The game loop cycle starts with initialising the scene and actors. Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate. This is illustrated by the loop in the physics update in Fig.~\ref{GameEngineLoopDiagram}. The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate~\cite{unity_framerates}. When the frame is rendered the game loop cycle returns to processing inputs. For an intuitive and  more detailed description of the interplay between the physics and render cycles, see~\cite{JohnAustinUnity}. 
% \footnote{\url{https://johnaustin.io/articles/2019/fix-your-unity-timestep}}.% is given

@@

% NEW TEXT V3
It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which necessitates rounding due to a fixed bit width~\cite{FloatingPointsBook,goldberg1991every}. 
% available and can potentially result in rounding errors~\cite{FloatingPointsBook,goldberg1991every}. 
As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and results may differ depending on execution order.
% Some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}. 
%
In the context of AV simulation, this
% finite-precision of floating point number representation 
% such rounding errors 
could result in accuracy issues of, for example, actor positions. While some authors suggest
avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. So, even if the result of a mathematical operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} when repeated for implementations that conform to the IEEE floating-point standard~\cite{8766229}. 
%
However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, 
parallelisation within the runtime environment or at the hardware level and performing the execution on a GPU rather than a CPU\footnote{ CPU and GPU processors may have different register widths~\cite{Whitehead2011} } may all affect the execution order. 
%
In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable, hardware, configuration and execution order.

@@

Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority depending on the operating system's scheduler policy. A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}. In principle, changing the scheduling policy and thread priorities may increase simulation variance.

It is responsibility of the tester
% It is unlikely 
It's unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should therefore give the same output. However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} 
% ~\cite{unity_multithreading} 
or if the scheduler simply randomly selects from a set of threads with equal priority, this may alter an otherwise deterministic sequence of events.

Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes. This may be decided based on factors such as throughput, latency or CPU utilisation. Scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, or executing across a parallel GPU architecture, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution. Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads, thus promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.

@@

The Unreal documentation and source code suggests that CARLA uses a priority queue based on a binary heap 
% e.g. see\footnote{\url{https://docs.unrealengine.com/en-US/API/Runtime/AIModule/FGraphAStar/index.html}, \url{https://github.com/EpicGames/UnrealEngine/blob/release/Engine/Source/Runtime/AIModule/Public/GraphAStar.h} and \url{https://docs.unrealengine.com/en-US/ProgrammingAndScripting/ProgrammingWithCPP/UnrealArchitecture/TArrays/index.html}}.
~\cite{FGraphAStar,TArray,GraphAStar}. 
Theoretically this means that use of A* in CARLA will be determinstic as long as the heap is always constructed in the same way.
However, if some unknown change causes the heap to be constructed in a different way, then the use of A* may then appear non-deterministic.
This can be a complex problem in practice.
For example, suppose A* iterates over actions in order of hash value, but between runs there is some untracked change to the hardware/software stack that alters the hash function, then this may change insertion order of nodes to the frontier, which could eventually lead to a different binary heap and thus a different optimal solution found by A*.
This suggests that while a simulated environment may behave deterministically, factors outside the simulator may cause changes to the operation of an A* implementation, which would then materialise as non-deterministic runs in the simulator.
The difference between use of a stable priority queue and an unstable but deterministic priority is simply that it may be easier to detect such changes under the former because insertion order is typically more meaningful with respect to the actual implementation of A*.

@@

Performing physics calculations and running the simulator may consume a significant proportion of system resources (e.g. CPU and GPU processors) for higher performing simulations. An initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance} which was corroborated with some initial investigations~\cite{TSLUnrealEngineTesting} and explored more fully in this work using the Carla platform.

@@

CPU utilisation software, such as the linux \texttt{stress} tool, which is a workload generator program, can be used to spawn workers on any number of cores or virtual threads on a system. This can be used to artificially increase the load on the system. For GPU utilisation, \texttt{gpu-burn} can be employed using the \texttt{fur test}. Different resolutions and multiple instances can be used to tune graphical utilisation levels~\cite{GPU_stress}. Reported values of resource utilisation can be obtained using the system monitors \texttt{htop} and \texttt{nvidia-smi} for CPU and GPU, respectively. These values should be added to the data logs. Alternatively, in place of artificial resource utilisation, multiple instances of the simulation could be executed simultaneously. However, the granularity of control with this approach may be reduced.

@@

\subsubsection{Scheduling}
We hypothesise that thread scheduling may be a major contributor to the non-deterministic results found in the empirical study. However, gaining fine control over the scheduling policy and thread execution order is non-trivial~\cite{acm-q-rr-interview}. The operating system schedules threads according to a specified scheduling policy, potentially based on equal thread priority. Thus, in such a case, all tasks non-essential to the simulator should be terminated to prevent interference with the simulation. Assigning a higher priority to the simulator process may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows\footnote{\url{https://docs.microsoft.com/en-us/windows/win32/taskschd/tasksettings-priority}}
% ~\cite{TaskSettingWindows} 
or \texttt{NICE} in Linux\footnote{\url{https://linux.die.net/man/1/nice}}.
%~\cite{Nice_linux}.

@@

An investigation into the CARLA simulator revealed a significant simulation variance for repeated tests with the same initial conditions and event history, indicating non-determinism of the simulation. We then researched, identified and discussed potential sources for non-determinism in this context. 
%
% A systematic case study of the CARLA simulator uncovered the actual factors that contribute towards greater simulation variance, giving rise to non-deterministic execution. 
%
In particular, actor collisions and system-level resource utilisation were identified as key contributors to increased simulation variance when using Carla and we recommend monitoring these during simulation execution or using a commercial alternative, e.g.~\cite{rFpro2021}. 
%
% To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution. 
%
% Alternatively, there are commercially available driving simulators that claim to be fully deterministic, e.g.\ RFPro~\cite{rFpro2021}, that may be more suitable if using a game engine does not provide a simulation variance sufficient for the verification requirements.

>>>>

\section{Preliminaries} \label{s:prelim}
\subsection{Definitions}
Refer to Fig.~\ref{variance_description} for the definitions introduced in this section.
% Several definitions are introduced in this section. These are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
\\

@@

A simulation can be thought of as the process of generating or producing experimental data. In the case of a driving simulator, kinematics will describe future states of actors given the current conditions and actions taken, thereby generating new data. If a simulation is deterministic, Fig.~\ref{variance_description_b}, then there will be no variation in the generated output data, i.e. all future states are perfectly reproducible from prior states and actions. However, if a simulation is non-deterministic, Fig.~\ref{variance_description_a}, then there will be a variation in the output data. \\

@@

Investigating the determinism of game engines has not attracted much research interest to date since performance is more critical for game developers than accurate and repeatable execution. Ensuring software operates deterministically is a non-trivial task.  Catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}. This section gives an overview of the internal structure of a game engine and what sources or settings may affect \textit{simulation variance}.

@@

The game loop cycle starts with initialising the scene and actors. Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate. This is illustrated by the loop in the physics update in Fig.~\ref{GameEngineLoopDiagram}. The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate~\cite{unity_framerates}. When the frame is rendered the game loop cycle returns to processing inputs. For an intuitive and  more detailed description of the interplay between the physics and render cycles see~\cite{JohnAustinUnity}. 
% \footnote{\url{https://johnaustin.io/articles/2019/fix-your-unity-timestep}}.% is given

@@

% NEW TEXT V3
It is a common misconception to attribute non-deterministic computational execution with the use of floating-point number representation, which necessitates rounding due to a fixed bit width~\cite{FloatingPointsBook,goldberg1991every}. 
% available and can potentially result in rounding errors~\cite{FloatingPointsBook,goldberg1991every}. 
As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and results may differ depending on execution order.
% Some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}. 
%
In the context of AV simulation, this
% finite-precision of floating point number representation 
% such rounding errors 
could result in accuracy issues of, for example, actor positions. While some authors suggest
avoiding floating-point representation entirely~\cite{empirical-analysis-of-flaky-tests}, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. So, even if the result of a mathematical operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} when repeated for implementations that meet the IEEE floating-point standard~\cite{8766229}. 
%
However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, 
parallelisation within the runtime environment or at the hardware level and performing the execution on a GPU rather than a CPU\footnote{ CPU and GPU processors may have different register widths~\cite{Whitehead2011} } may all affect the execution order. 
%
In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable, hardware, configuration and execution order.

@@

Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority depending on the operating system's scheduler policy. A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}. In principle, changing the scheduling policy and thread priorities may increase simulation variance. 
%
It would therefore be important to ensure these remain stable between repeated runs. 
% Changes to thread priorities or scheduling policy during repeated runs on the same hardware and should therefore give the same output. 
However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} 
% ~\cite{unity_multithreading} 
or if the scheduler simply randomly selects from a set of threads with equal priority, this may alter an otherwise deterministic sequence of events.

Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes. This may be decided based on factors such as throughput, latency or CPU utilisation. Scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution. Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads, thus promoting deterministic behaviour by design~\cite{acm-q-rr-interview}. 
%
Likewise, simulation variance may be observed for game engines that use GPU parallelisation  by offloading time-critical calculations to several dedicated cores simultaneously. While this would be faster than a serial execution, the order of execution arising from program-level concurrency cannot, in general, be guaranteed.

@@

The Unreal documentation and source code suggest that CARLA uses a priority queue based on a binary heap 
% e.g. see\footnote{\url{https://docs.unrealengine.com/en-US/API/Runtime/AIModule/FGraphAStar/index.html}, \url{https://github.com/EpicGames/UnrealEngine/blob/release/Engine/Source/Runtime/AIModule/Public/GraphAStar.h} and \url{https://docs.unrealengine.com/en-US/ProgrammingAndScripting/ProgrammingWithCPP/UnrealArchitecture/TArrays/index.html}}.
~\cite{FGraphAStar,TArray,GraphAStar}. 
Theoretically this means that use of A* in CARLA will be deterministic as long as the heap is always constructed in the same way.
However, if some unknown change causes the heap to be constructed in a different way, then the use of A* may appear non-deterministic.
This can be a complex problem in practice.
For example, suppose A* iterates over actions in order of hash value, but between runs there is some untracked change to the hardware/software stack that alters the hash function, then this may change insertion order of nodes to the frontier, which could eventually lead to a different binary heap and thus a different optimal solution found by A*.
This suggests that while a simulated environment may behave deterministically, factors outside the simulator may cause changes to the operation of an A* implementation, which would then materialise as non-deterministic runs in the simulator.
The difference between use of a stable priority queue and an unstable but deterministic priority is simply that it may be easier to detect such changes under the former because insertion order is typically more meaningful with respect to the actual implementation of A*.

@@

Executing physics calculations during simulation may consume a significant proportion of system resources (e.g.\ CPU and GPU processors). This suggests that as resource utilisation of the simulation increases, so does the \textit{simulation variance} which was corroborated with some initial investigations~\cite{TSLUnrealEngineTesting} and explored more fully in this work using CARLA.

@@

CPU utilisation software, such as the linux workload generator \texttt{stress}, %which is a workload generator program, 
can be used to spawn workers on any number of cores or virtual threads on a system. This can be used to artificially increase the load on the system. For GPU utilisation, \texttt{gpu-burn} can be employed using the \texttt{fur test}. Different resolutions and multiple instances can be used to tune graphical utilisation levels~\cite{GPU_stress}. Reported values of resource utilisation can be obtained using the system monitors \texttt{htop} and \texttt{nvidia-smi} for CPU and GPU, respectively. These values should be added to the data logs. Alternatively, in place of artificial resource utilisation, multiple instances of the simulation could be executed simultaneously. However, the granularity of control with this approach may be reduced.

@@

\subsubsection{Scheduling}
We hypothesise that thread scheduling may be a major contributor to the non-deterministic results of the empirical study. However, gaining fine control over the scheduling policy and thread execution order is non-trivial~\cite{acm-q-rr-interview}. The operating system schedules threads according to a specified scheduling policy, potentially based on equal thread priority. Thus, in such a case, all tasks non-essential to the simulator should be terminated to prevent interference with the simulation. Assigning a higher priority to the simulator process may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows\footnote{\url{https://docs.microsoft.com/en-us/windows/win32/taskschd/tasksettings-priority}}
% ~\cite{TaskSettingWindows} 
or \texttt{NICE} in Linux\footnote{\url{https://linux.die.net/man/1/nice}}.
%~\cite{Nice_linux}.

@@

An investigation into the CARLA simulator revealed a significant simulation variance for repeated tests with the same initial conditions and event history, indicating non-determinism of the simulation. We then researched, identified and discussed potential sources for non-determinism in this context. 
%
% A systematic case study of the CARLA simulator uncovered the actual factors that contribute towards greater simulation variance, giving rise to non-deterministic execution. 
%
In particular, actor collisions and system-level resource utilisation were identified as key contributors to increased simulation variance when using CARLA and we recommend monitoring these during simulation. Alternatively some commercial driving simulators claim to be fully deterministic, for instance RFpro~\cite{rFpro2021}. 
%
% To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution. 
%
% Alternatively, there are commercially available driving simulators that claim to be fully deterministic, e.g.\ RFPro~\cite{rFpro2021}, 
These may be more suitable if using a game engine does not provide a simulation variance sufficient for the verification requirements.
